---
title: 'Google Data Analytics Capstone Project: Case Study 1'
author: "Irfan Peterson"
date: "2022-08-15"
output:
  html_document: default
  pdf_document: default
---


## Clearing Workspace
```{r}
rm(list=ls())
```

## Setting Workspace
```{r}
setwd("~/Google Data Analytics/8. Google Data Analytics Capstone,  Complete a Case Study/irfan_peterson_case_study_1")
```


## Setting Up My Environment by Installing and Loading Various Analysis Packages

```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("here", repos = "http://cran.us.r-project.org")
install.packages("skimr", repos = "http://cran.us.r-project.org")
install.packages("janitor", repos = "http://cran.us.r-project.org")
install.packages("DescTools", repos = "http://cran.us.r-project.org")


library("tidyverse") # A system of packages for data manipulation, exploration and visualization
library("dplyr") # Useful for common data manipulation tasks
library("ggplot2") # Useful for creating various visualizations
library("readr") # Tool for reading rectangular data, useful for importing csv dataset
library("tidyr") # Tool for data cleaning
library("lubridate") # Used to convert different data types into date and time formats
library("here") # Makes referencing easier
library("skimr") # Useful to make summarizing data easier
library("janitor") # Has functions for cleaning data
library("DescTools") #Tools for descriptive analysis
```



## Importing Cyclist Datasets and Assigning Them to a More Readable Variable

```{r}
td_08_21 <- read_csv("202108-divvy-tripdata.csv")
td_09_21 <- read_csv("202109-divvy-tripdata.csv")
td_10_21 <- read_csv("202110-divvy-tripdata.csv")
td_11_21 <- read_csv("202111-divvy-tripdata.csv")
td_12_21 <- read_csv("202112-divvy-tripdata.csv")
td_01_22 <- read_csv("202201-divvy-tripdata.csv")
td_02_22 <- read_csv("202202-divvy-tripdata.csv")
td_03_22 <- read_csv("202203-divvy-tripdata.csv")
td_04_22 <- read_csv("202204-divvy-tripdata.csv")
td_05_22 <- read_csv("202205-divvy-tripdata.csv")
td_06_22 <- read_csv("202206-divvy-tripdata.csv")
td_07_22 <- read_csv("202207-divvy-tripdata.csv")
```

## Reviewing the Data Frame of the Dataset Using Glimpse Function
```{r}
glimpse(td_08_21)
glimpse(td_09_21)
glimpse(td_10_21)
glimpse(td_11_21)
glimpse(td_12_21)
glimpse(td_01_22)
glimpse(td_02_22)
glimpse(td_03_22)
glimpse(td_04_22)
glimpse(td_05_22)
glimpse(td_06_22)
glimpse(td_07_22)
```


## Combining the Twelve Datasets into One Dataset
```{r}
td_all <- rbind(td_01_22,td_02_22, td_03_22, td_04_22, td_05_22, td_06_22, td_07_22, td_08_21, td_09_21, td_10_21, td_11_21, td_12_21)
```

## Formatting and Verifying the Structure of the Combined Dataset
```{r}

td_all$member_casual <- as.factor(td_all$member_casual) # Formatting membership column to format for analysis
clean_names(td_all) # Clean column names and make the formatting consistent

str(td_all) # Reviewing the structure of the combined data set
```


## Introducing New Variables for Analysis
```{r}
nv_td_all <- td_all %>%
  mutate(trip_length_sec = ended_at - started_at) %>% #Compute duration of trip
  mutate(trip_length_hms = seconds_to_period(trip_length_sec)) #%>% #Duration in HH:MM:SS for better readability for users
  
nv_td_all$start_day <- weekdays(td_all$started_at, abbreviate = TRUE) #Output the day when the trip started
nv_td_all$start_day <- ordered(nv_td_all$start_day, levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) #ordering days of the week

nv_td_all$sd_yr <- year(nv_td_all$started_at) # Numeric column for year
nv_td_all$sd_mnth <- month(nv_td_all$started_at) # Numeric column for month
nv_td_all$sd_dy <- day(nv_td_all$started_at) # Numeric column for day
nv_td_all$sd_hr <- hour(nv_td_all$started_at) # Numeric column for hour

```


## Cleaning the Dataset
```{r}

nv_td_all <- subset(nv_td_all, trip_length_sec > 0) # Take out negative values from the new variable, trip_length_sec
q1 <- quantile(nv_td_all$trip_length_sec, .25) # Establishes the 25th quartile value
q3 <- quantile(nv_td_all$trip_length_sec, .75) # Establishes the 75th quartile value
iqr <- IQR(nv_td_all$trip_length_sec)
no_outl_td <- subset(nv_td_all, nv_td_all$trip_length_sec > (q1 - 1.5 * iqr) & nv_td_all$trip_length_sec < (q3 + 1.5 * iqr)) #Removes outliers from the dataset


cln_td_all <- no_outl_td %>%
  drop_na() %>% # Removes NA values
  distinct() # %>% # Removes duplicate data
  #str_trim() # Removes trailing and leading white spaces

```
 
 

## Descriptive & Regression Analysis
```{r}
# Measure of central tendency (max, min, mode,mode)
aggregate(trip_length_sec ~ member_casual, cln_td_all, mean) # Mean  of trip length
aggregate(trip_length_sec ~ member_casual, cln_td_all, max) # Max of trip length
aggregate(trip_length_sec ~ member_casual, cln_td_all, min) # Min of trip length
Mode(cln_td_all$start_day) # Finding the most frequent day bikes are being rented out

# Average length of ride between members and casual riders
cln_td_all %>%  
  group_by(member_casual) %>%
  summarize(avg_tl = mean(trip_length_sec))

# Average length of ride for users by day of the week
aggregate(trip_length_sec ~ start_day + member_casual, cln_td_all, mean)

# Logistic regression to predict member & casual riders
log_reg <- glm(member_casual ~ start_day, family = "binomial", data = cln_td_all)
summary(log_reg)

```

## Exporting the Data as CVS File
```{r}
write.csv(cln_td_all, "cln_case_study_1")
```

## Data Visualizations
```{r}

#1. Bar chart of bike rented per day, separated between member and casual rider
ggplot(cln_td_all)+
  geom_bar(mapping = aes(x = start_day, fill = start_day))+
  labs(title = "Frequency of Bike Rented by Day", subtitle = "Over the Past 12 Months", x = "Day of the Week")+
  facet_wrap(~member_casual)

#2. Bar chart of distribution of type of bikes, separated between member and casual rider
ggplot(cln_td_all)+
  geom_bar(mapping = aes(x = rideable_type, fill = rideable_type))+
  labs(title = "Frequency of Bike Rented by Type of Bike", x = "Classification of Bike")+
  guides(x = guide_axis(n.dodge = 2))+ 
  facet_wrap(~member_casual)

#3. Bar chart of distribution of bikes rented, grouped by days of the week between member and casual rider
ggplot(cln_td_all, aes(x = member_casual, y = trip_length_sec, fill = member_casual))+
  geom_bar(stat = "identity")+
  labs(title = "Count of Bike Rented Seperated by Day of the Week", x = "Membership Status")+
  guides(x = guide_axis(n.dodge = 2))+ 
  facet_grid(~start_day)+
  scale_fill_manual(values = c("casual" = "cornflowerblue",
                               "member" = "chocolate1"))

#4. Bar chart of distribution of bikes rented, grouped by month between member and casual rider
ggplot(cln_td_all)+
  geom_bar(mapping = aes(x = member_casual, fill = member_casual))+
  labs(title = "Frequency of Bike Rented by Month", x = "Membership Status")+
  guides(x = guide_axis(n.dodge = 2))+ 
  facet_grid(~sd_mnth)+
  scale_fill_manual(values = c("casual" = "cornflowerblue",
                               "member" = "chocolate1"))

#5. Bar hart of distribution of bikes rented, rented by Hour of the day, separated by day of the Week and membership
ggplot(cln_td_all)+
  geom_bar(mapping = aes(x = sd_hr, fill = member_casual))+
  labs(title = "Distribution of Bike Rented, Seperated by Hour, Membership and Day of the Week", x = "Hour of the Day")+
  guides(x = guide_axis(n.dodge = 2))+ 
  facet_grid(start_day~member_casual)+
  scale_fill_manual(values = c("casual" = "cornflowerblue",
                               "member" = "chocolate1"))

```

## Thank you!!!